### 2018-07-02 paper/5-implementations.tex
Work on S5

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key value (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) (@@)
    build key = case tasks key of
        Nothing -> return (@@)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do lift (build k)                      -- build the key
                                 lift (gets (getInfo . fst)) >>= put -- save new traces
                                 lift (gets (getValue k . fst))      -- fetch the value
                newValue <- flattenInfo (run newTask fetch)
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Collapse two layers of stateful computations into one (implementation omitted)
flattenInfo :: StateT i (State (Store i k v, Set k)) a -> State (Store i k v, Set k) a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-6mm}
\end{figure}

@@

These steps are run in the
\hs{StateT}~\hs{i}~\hs{(@@State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k))}
monad, because the \hs{newTask} has access only to the~\hs{i} part of the state;
we therefore need to use \hs{lift} when executing commands in the inner
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
The pair \hs{(@@newValue,}~\hs{newInfo)} obtained by running the \hs{newTask} is
stored, and the \hs{key} is added to the set \hs{done}.

\subsection{Cloud build systems: \Bazel, \CloudBuild and \Cloud \Shake}\label{sec-implementation-cloud}

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    if value `elem` cachedValues then return value else case cachedValues of
        (cachedValue:_) -> return cachedValue
        _ -> do (newValue, deps) <- trackM task fetch
                put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
                return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to ctRebuilder
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
chain.}
\label{fig-cloud-implementations}
\vspace{-4mm}
\end{figure}

In this section we use constructive traces to implement cloud build systems.
Fig.~\ref{fig-cloud-implementations} shows our models of \Bazel,
\CloudBuild and \Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a
rebuilder based on constructive traces (\S\ref{sec-constructive-traces}).

The implementation of \hs{ctRebuilder} is analogous to that of \hs{vtRebuilder}
in Fig.~\ref{fig-shake-implementation}, but the \hs{verifyVT} query is replaced
with a more powerful query to \hs{constructCT} that can construct a resulting
\hs{cachedValue} by looking it up the cloud cache. If the cache lookup fails, we
fall back to executing the \hs{task}. The obtained \hs{newValue} and the
\hs{task}'s dependencies are recorded as a new constructive trace for future
use.

The \Bazel build system uses a restarting scheduler whose implementation we
omit. It is similar to \Excel's \hs{restarting} scheduler defined in
Fig.~\ref{fig-excel-implementation}, but instead of building keys in the order
specified by the persistently stored calculation chain, \Bazel uses a
\emph{build queue}. The build starts with the queue containing all dirty keys.
Similar to \Excel, the rebuilding of a key extracted from the queue may fail
because one of its dynamic dependencies is dirty. In this case the key is marked
as \emph{blocked} and its rebuilding is deferred. Whenever a key is successfully
rebuilt, all keys that were previously blocked on it are added back to the
queue, and their build is eventually restarted.

@@

\textbf{TODO:} A real implementation of \Cloud \Shake is now available... It was
obtained by directly applying the presented ideas.

\subsection{Buck and Nix}

Some build systems allow to exploit \emph{build task determinism} for efficient
fetching of build results from the cache: one can simply hash the values of all
transitive inputs of a key (which can be done efficiently by storing all hashes
in a Merkle tree) and download the matching value from the cache, without the
need to build any intermediate values. This can bring significant performance
benefits in practice, but sacrifices the early cutoff property: indeed, with
this approach addition of a new comment to one of the source files will
invalidate all dependent keys.

Fig.~\ref{fig-cloud-implementations} shows our models of \Buck and \Nix build
systems that use the \hs{dctRebuilder} -- a rebuilder based on deep
constructive traces (\S\ref{sec-deep-constructive-traces}). As one can
see, \hs{dctRebuilder} is almost identical to \hs{ctRebuilder}, but it ignores
the current \hs{value}, since the \hs{constructDCT} query only needs to examine
the transitive inputs of the given \hs{key}. \Buck uses the \hs{topological}
scheduler and is therefore an applicative build system, whereas \Nix uses the
\hs{suspending} scheduler and is monadic.

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
        return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) (@@)
    build key = case tasks key of
        Nothing -> return (@@)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do lift (build k)                      -- build the key
                                 lift (gets (getInfo . fst)) >>= put -- save new traces
                                 lift (gets (getValue k . fst))      -- fetch the value
                newValue <- flattenInfo (run newTask fetch)
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Collapse two layers of stateful computations into one (implementation omitted)
flattenInfo :: StateT i (State (Store i k v, Set k)) a -> State (Store i k v, Set k) a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-6mm}
\end{figure}

@@

These steps are run in the
\hs{StateT}~\hs{i}~\hs{(@@State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k))}
monad, because the \hs{newTask} has access only to the~\hs{i} part of the state;
we therefore need to use \hs{lift} when executing commands in the inner
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
We use the helper function \hs{flattenInfo} to collapse two layers of stateful
computations into one. The \hs{newValue} obtained by running the \hs{newTask} is
stored, and the \hs{key} is added to the set \hs{done}.

\subsection{Cloud build systems: \Bazel, \CloudBuild, \Cloud \Shake, \Buck and \Nix}\label{sec-implementation-cloud}

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    if value `elem` cachedValues then return value else case cachedValues of
        (cachedValue:_) -> return cachedValue
        _ -> do (newValue, deps) <- track task fetch
                modify $ recordCT key newValue [ (k, hash v) | (k, v) <- deps ]
                return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to ctRebuilder
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
chain.}
\label{fig-cloud-implementations}
\vspace{-4mm}
\end{figure}

Fig.~\ref{fig-cloud-implementations} shows our models of several cloud build
systems. \Bazel, \CloudBuild and \Cloud \Shake are based on constructive traces
(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
version (\S\ref{sec-deep-constructive-traces}).

The implementation of \hs{ctRebuilder} is analogous to that of \hs{vtRebuilder}
in Fig.~\ref{fig-shake-implementation}, but the \hs{verifyVT} query is replaced
with a more powerful query to \hs{constructCT} that returns a list of suitable
\hs{cachedValues} by looking them up the cloud cache. If the current \hs{value}
is in the list, we can use it as is. Otherwise, if the list is non-empty, we can
use an arbitrary \hs{cachedValue}. Finally, if the cache has no suitable values,
we fall back to executing the \hs{task}. The obtained \hs{newValue} and the
\hs{task}'s dependencies are recorded as a new constructive trace for future
use.

The \Bazel build system uses a restarting scheduler whose implementation we
omit. It is similar to \Excel's \hs{restarting} scheduler defined in
Fig.~\ref{fig-excel-implementation}, but instead of building keys in the order
specified by the persistently stored calc chain, \Bazel uses a \emph{build
queue}. The build starts with the queue containing all dirty keys. Similar to
\Excel, the rebuilding of a key extracted from the queue may fail because one of
its dynamic dependencies is dirty. In this case the key is marked as
\emph{blocked} and its rebuilding is deferred. Whenever a key is successfully
rebuilt, all keys that were previously blocked on it are added back to the
queue, and their build is eventually restarted.

@@

Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the
\hs{dctRebuilder}, a rebuilder based on deep constructive traces
(\S\ref{sec-deep-constructive-traces}), whose implementation we omit since it
is very similar to that of \hs{ctRebuilder}. \Buck uses the \hs{topological}
scheduler and is therefore an applicative build system, whereas \Nix uses the
\hs{suspending} scheduler and is monadic. Deep constructive traces allow these
build systems to efficiently support shallow builds: the value of the target key
can be downloaded by examining only the hashes of all transitive inputs,
skipping the intermediate keys. The downside of this approach is the loss of the
early cutoff property.

% Some build systems allow to exploit \emph{build task determinism} for efficient
% fetching of build results from the cache: one can simply hash the values of all
% transitive inputs of a key (which can be done efficiently by storing all hashes
% in a Merkle tree) and download the matching value from the cache, without the
% need to build any intermediate values. This can bring significant performance
% benefits in practice, but sacrifices the early cutoff property: indeed, with
% this approach addition of a new comment to one of the source files will
% invalidate all dependent keys.
